\modLP{Identification of} Biological Regulatory Networks from Process Hitting models

### 2014-09-03 response.tex

\documentclass[11pt]{article}
% vim:set spell spelllang=en:
\usepackage{color}
\usepackage{fullpage}

% ----- parameters ----------------------------------------------------------------
\def\BibTex{{\rm B\kern-.05em{\sc i\kern-.025em b}\kern-.08em
         T\kern-.1667em\lower.7ex\hbox{E}\kern-.125emX}}
%\def\baselinestretch{1.5}
\usepackage{amsmath}
\usepackage{amssymb}

% to import paper citations and refs
% generated with 
% cat ../v5/paper.aux  | sed 's/\\@/% \\@/' > paper.aux
%\input{paper.aux}

% ----------------------------------------------------------------------------------
\newcounter{pointno}
\setcounter{pointno}{0}
\def\thepoint{\arabic{pointno}}
\newcommand{\ptno}[1]{\textcolor{blue}{\refstepcounter{pointno} \thepoint. #1}}
\newcommand{\question}[1]{\bigskip{\it\ptno{#1}}}
\def\answer{}

\def\TODO#1{\textbf{[TODO: #1]}}
\def\DISCUSS#1{\textbf{DISCUSSION(#1): }}

\setlength{\parindent}{0pt}

\def\V{\mathcal V}
\def\O{\mathcal O}

\def\zero{\mathbf 0}
\def\R{\mathbb R}
\def\Rpos{\R_{\geq 0}}
\def\Rspos{\R_{>0}}
\def\Q{\mathbb Q}
\def\Qpos{\Q_{\geq 0}}
\def\Qspos{\Q_{>0}}
\def\Z{\mathbb Z}
\def\Zpos{\Z_{\geq 0}}
\def\Zspos{\Z_{> 0}}

\def\mspan{\mathrm{span}}
\def\lp{\mathrm{lowerpoint}}
\def\orderings{\mathrm{orderings}}

\def\inner#1#2{#1#2}
%\def\inner#1#2{\langle #1,#2 \rangle}

\def\apply{\bullet}

\def\B{\mathcal B}

\begin{document}

\section*{Comments and answers to the reviewer}

We thank the reviewer for th comments and  suggestions.

\TODO{summary of response + changes}

Please find below our detailed answers.

\question{Beside, the definition of RT model, including signs that can be
deduced
from the K parameters must be justified or clarify. For a computer scientist
it is always surprising to have redundant pieces of information without accounting their matching.
With your definition I can write $a \xrightarrow+ b$ and I can
set $K_{b,\emptyset} = 1$ and $K_{b,\{a\}} = 0$.
On the one side I define a as an activator and
on the other side I define a as an inhibitor with K parameters. As signs do
not enter in the interpretation, why they are included in the formalism as
necessary ``comments''?}

\DISCUSS{LP}
We have to clarify this point, while insisting that we want to stick with this
standard definition (due to Thomas). Arguments:
\begin{enumerate}
\item this design allows to decouple the IG from the dynamics, which is
interesting for modelling biological systems as the IG is often a prior
knowledge with possible inconsistencies with the dynamics we want to encode
with.
A further step may then infer back the IG of the actual dynamics (here, an
abstraction of the encoded dynamics) and compare it with the prior IG. Typically, if the inferred IG is included in the prior
one, we have consistency.
\item Additional constraints (as discussed in sect XX) on parameters can also
ensure consistency between the prior IG and the encoded dynamics.
For instance, monotonicity (which imposes an ordering between the value of
parameters), but also lighter constraints (like imposing minimum and maximum
values).
Hence, the K definition is very generic, ranging from totally decoupled
specification to very constrained valuation w.r.t to the prior IG.
\item Several alternative definition exist, with various degree of equivalency.
We picked this definition as it the original and standard one in the Thomas
community (give reference to Thomas' articles).
\end{enumerate}

\TODO{Answer inc. text modification}

\question{
(P.1) Abstract: the following sentence is not understandable. Either more explanations
are needed or simply suppress it.
``Its major originality lies in a specific restriction on the causality of actions, which makes the formal
analysis of very large systems tractable.''
}

\answer
\DISCUSS{LP}
proposition: ``The PH is a particular class of Indeterministic Asynchronous Automata
Network (or safe Petri nets) on which have been designed static analysis for
dynamical properties that are tractable on very large systems.''

\TODO{Modify sentence + answer (quote)}

\question{
The BRNs were introduced by S. Kauffman in 1969. His work also covers
“numerous theoretical works”. Therefore, it would be interesting to mention this
fact. Ren\'e Thomas is not the father of BRN applied to regulatory
circuit dynamics. Beside, the definition of the BRN for S. Kaufman consists
of a Boolean system dynamics with a Boolean function which is a normal
way to define a dynamical system. The R.T.'s formalization is slightly unconventional, namely not based on a stepwise function.}

\answer
\DISCUSS{LP} yes, we have to mention Kauffman of course, while stressing the
differences brought by Thomas (async + parameters).

\question{
For Boolean networks, symbolic methods and parsimonious generation of
states exist to circumvent the generation of the whole state graph and then
scaling up the performances of the analysis (e.g. equilibria computation).
Therefore, claiming that it is a ``major difficulty'' is partly true only.
}

\answer
\DISCUSS{LP} yes, we have to mention those work, while stressing that they have
hard time with transient dynamics on very large networks, partly due to
concurrency, which makes model-checking very hard even with symbolic techniques.

\DISCUSS{LP} how such a discussion is relevant for the paper? Can't we simply
say that PH allows to model indeterministic Boolean and multi-valued networks
and remove (or drastically reduce) this discussion on model-checking?

\question{
(P.3 and 6) The term “cooperation” is often mentioned. It is not self-
explanatory for the reader. In particular in P.3 “associated influences” is
not explanatory at all. If I have missed it I suggest to more clearly emphasize it. This comment does not concern cooperative sort which is clearly
explained, but the notion of cooperation that cooperative sort encompasses.
A definition of “cooperation” is required (formal or not) or more explanations of the notion of cooperation in this context.
}

\DISCUSS{LP} we should avoid this term in introduction, and only use once
formally defined.
In the introduction, maybe we should mention that
PH models can embed additional automata that encode for some (partial) logic
gates. Such additional automata that can be considered as ``internal''
variables, and we may want to mask them in abstract representations of the
systems, such as IG and Thomas' parameters.

\question{
(P. 6) The arrows representing the hitting of sort a fig 1. are too small.
}

\answer
\TODO{improve figure}

\question{
(P. 8)
``This allows the existence of interleaving of actions leading .. semantics with
prioritized actions defined in [13]''.
Again I don't understand the message related to your own notion of cooperation. Is it really important to
mention these elements?}

\answer
\DISCUSS{LP} I would go for removing this discussion.

\question{
(P. 9) Interaction Graph. $\Gamma$ refers to the set of vertices. However
$\Gamma^{−1}(b)$
requires to seek the pre-image in E. I find the choice of $\Gamma$ to annotate
a function giving the regulators confusing. If we admit that E defines a
relation, $Eb$ or $E^{-1} (b)$ are more natural notations for the pre-image or more
simply $b \rightarrow$ for the image and $\rightarrow b$ for the pre image. These notations
are classically used for relation to qualify these sets of elements and do not
require additional notations.
}

\answer
\TODO{Apply the changes + quick answer}

\question{
(P. 8) You mention that Thomas formalism lies on two complementary descriptions of the system IG and K parameters. However There exists some
redundancies for these description because the sign may be deduced from the
parameters representing monotonous properties of the parameters. Moreover, The dynamics is not deduced from signs. IG graph with levels and
parameters appears to suffice for fully describing the RT model. It would
be more parsimonious (for an article in TCS) to define an IG with levels
only, and parameters corresponding to Thomas’ model and then explain
that signs are abstractions of the parameters corresponding to monotony
properties.
}

\answer
\DISCUSS{LP} See 1. The definitions need to be more clear with more textual
explanation for CS people.

\question{
(P. 11) Definition of the sign. The term of monotony is classically associated
to the inference of signs and should be clearly mentioned. Here it is hidden
in your definition. The discussion on the number of maximum levels has
already been fully developed in other articles related to RT formalisms. Is
it necessary here just for mentioning that PH may accept any level?
}

\answer
\DISCUSS{LP} Yes, we should make clear that we are looking for monotonic
influences - but not necessarily at the global scope, which mean that we can
find two contradictory monotonic influences, leading to those ``unsigned'' edges
in the IG.
And yes, we should just mention that PH models can have an arbitrary number of
levels.

\question{
(P. 12) The order of your explanations makes the text confusing. Signs
are components of the definition of an IG. After you argue that it is useless
because parameters suffice to deduce the state graph. You mention that they
can be inferred although they must be included in the definition of an IG.
The formalization is not clear. Unsigned IG could/must be presented before
signed IG which is an abstraction of the monotony of the K parameters.
}

\answer
\DISCUSS{LP} Same as 1 and 9.

\question{
(P. 12) ``However, it is common that the prior IG actually refers to interaction
.. enhancing the conclusiveness of static analyses''.
The inference of an IG from a dynamics is interesting and important in biology. However the
motivations expressed here are not well argued. In particular I don't find
this "common". Some bibliographical references are required to consolidate
the arguments or the sentence has to be rephrased.
}

\answer
\DISCUSS{LP}
Indeed, the motivation is quite weak. Some arguments:
\begin{enumerate}
\item Abstract representation of influences
\item Consistency with prior knowledge: if the modelling started from an IG, we may want to
verify that the dynamical models is consistent w.r.t. to the prior IG
\item + remove non-used influences (IG reduction).
\end{enumerate}

\question{
(P. 13 -14 - 15) 3 pages contain 5 definitions with no example and no
intuition on their use but for $\Gamma$ and focal. More explanations are required:
what kinds of concept do represent focal, what is the importance of a strict
context, where is the outline guiding the reader through the dense area of
definitions located in three pages only? These pages must be rewritten and
extended to understand the core of the method.
}

\answer
\DISCUSS{LP}redo this section

\question{
(P. 18) Figure 5 shows a result of Fig 3. However the detail concerns example
in Fig. 1?  Examples and the implementation part are more clear. However, the
biological example is only used to evaluate performances of the implementation. A concrete (biological) modeling example used as a guideline for
illustrating the different notions throughout the article would certainly help
the reader.
}

\answer
\DISCUSS{LP} use Lambda phage as a running example (or dedicate a subsection to
apply the whole pipeline to it).

\end{document}

### 2014-09-03 main.tex
intro

\documentclass[preprint]{elsarticle}

@@

@@

\title{Constructing Biological Regulatory Networks from Process Hitting models}
%\\\rewriteil{Or possibly: “Complementarity between PH, IG and Thomas modeling”}

\author[irccyn,nii]{Maxime Folschette}
\ead{Maxime.Folschette@irccyn.ec-nantes.fr}
\author[eth]{Lo\"ic Paulev\'e}
\author[nii]{Katsumi Inoue}
\author[irccyn,nii]{Morgan Magnin}
\author[irccyn]{Olivier Roux}

\address[irccyn]{LUNAM Universit\'e, \'Ecole Centrale de Nantes, IRCCyN UMR CNRS 6597\\
(Institut de Recherche en Communications et Cybern\'etique de Nantes)\\
1 rue de la No\"e - B.P. 92101 - 44321 Nantes Cedex 3, France.}
\address[nii]{National Institute of Informatics,\\
2-1-2, Hitotsubashi, Chiyoda-ku, Tokyo 101-8430, Japan.}
\address[eth]{BISON group, Automatic Control Laboratory, ETH Zürich\\
Physikstrasse 3, 8092 Zurich, Switzerland.}

\begin{abstract}
\parindent 0.5cm
The Process Hitting (PH) is a recently introduced framework to model concurrent processes.
\todo{The reviewer did not understand the following sentence:}
Its major originality lies in a specific restriction on the causality of actions, which
makes the formal analysis of very large systems tractable.
PH is suitable to model Biological Regulatory Networks (BRNs) with complete or partial
knowledge of cooperations between regulators by defining the most permissive dynamics
with respect to these constraints.

\todo{Rather name Kaufman as the creator of BRNs.}
On the other hand, the qualitative modeling of BRNs has been widely addressed using Ren\'e Thomas'
formalism, leading to numerous theoretical work and practical tools to understand emerging behaviors.
%In addition to be able to use the existing results on this formalism,
%obtaining the Thomas representation of a PH model allows to iteratively refine a model.
Linking this representation to PH would not only allow to use the existing results on this formalism,
but also to iteratively refine the model of a system.

This paper establishes formal links between these two formalisms, permitting to appreciate their
complementary features.
First, the inference of the Interaction Graph (IG) from a PH model gives a compact representation of
the regulations that are effective for the dynamics.
From this IG, well-known results can be applied to derive global dynamical properties for the
modeled BRN.
Second, the inference of all Thomas models that are compatible with a given PH is tackled.
As the PH allows partial specification of cooperation between components, this inference
emphasizes the ability of PH to deal with large BRNs with incomplete knowledge on
cooperations, where Thomas' approach fails because of the combinatorics of parameters.

The inference of corresponding Thomas' models is provided using Answer Set Programming,
which allows notably an efficient enumeration of (possibly numerous) compatible parametrizations.
\end{abstract}

>>>>

\documentclass[preprint]{elsarticle}
% vi:spell spelllang=en

@@

\def\modMF#1{\textcolor{teal}{#1}}
\def\modLP#1{\textcolor{magenta}{#1}}
\def\modMM#1{\textcolor{blue}{#1}}
\def\modOR#1{\textcolor{olive}{#1}}

@@

\title{\modLP{Identification of} Biological Regulatory Networks from Process Hitting models}
%\\\rewriteil{Or possibly: “Complementarity between PH, IG and Thomas modeling”}

\author[irccyn,nii]{Maxime Folschette}
\ead{Maxime.Folschette@irccyn.ec-nantes.fr}
\author[lri]{Lo\"ic Paulev\'e}
\author[nii]{Katsumi Inoue}
\author[irccyn,nii]{Morgan Magnin}
\author[irccyn]{Olivier Roux}

\address[irccyn]{LUNAM Universit\'e, \'Ecole Centrale de Nantes, IRCCyN UMR CNRS 6597\\
(Institut de Recherche en Communications et Cybern\'etique de Nantes)\\
1 rue de la No\"e - B.P. 92101 - 44321 Nantes Cedex 3, France.}
\address[nii]{National Institute of Informatics,\\
2-1-2, Hitotsubashi, Chiyoda-ku, Tokyo 101-8430, Japan.}
\address[lri]{CNRS, Laboratoire de Recherche en Informatique UMR CNRS 8623\\
Universit\'e Paris-Sud, 91405 Orsay Cedex, France}

\begin{abstract}
\modMF{Maxime}
\modLP{Lo\"ic}
\modMM{Morgan}
\modOR{Olivier}

%\parindent 0.5cm
The Process Hitting (PH) \modLP{is a particular class of Indeterministic Asynchronous Automata
Network (or safe Petri nets) on which have been recently designed several static analyses for
dynamical properties that are tractable on very large systems.}

\modLP{In this paper, we address the formal identification of qualitative
models of Biological Regulatory Networks (BRNs) from PH models.}
First, the inference of the Interaction Graph (IG) from a PH model
\modLP{summarizes the signed influences between the components} that are effective for the dynamics.
Second, we provide the inference of all Ren\'e Thomas models of BRNs 
that are compatible with a given PH.
As the PH allows the specification of indeterministic interactions between
components, our inference emphasizes the ability of PH to deal with large BRNs with incomplete knowledge
on interactions, where Thomas' approach fails because of the combinatorics of parameters.

The inference of corresponding Thomas models is provided using Answer Set Programming,
which allows notably an efficient enumeration of (possibly numerous) compatible parametrizations.

%\todo{Rather name Kaufman as the creator of BRNs.}
%On the other hand, the qualitative modeling of BRNs has been widely addressed using Ren\'e Thomas'
%formalism, leading to numerous theoretical work and practical tools to understand emerging behaviors.
%In addition to be able to use the existing results on this formalism,
%obtaining the Thomas representation of a PH model allows to iteratively refine a model.
%Linking this representation to PH would not only allow to use the existing results on this formalism,
%but also to iteratively refine the model of a system.

%This paper establishes formal links between these two formalisms, permitting to appreciate their
%complementary features.
\end{abstract}
